---
title: "Retail Sales Performance Analysis Using SQL and R"
author: "LaToya Burdette"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 5
)

library(tidyverse)
library(readxl)
library(janitor)
library(DBI)
library(RSQLite)
library(lubridate)
library(scales)

theme_set(theme_light())
```
## Overview

This project looks at retail sales data using SQL and R. The goal was to better understand revenue trends across customers, products, categories, suppliers, and regions. I imported the data from Excel, loaded it into a SQLite database, queried it with SQL, and visualized the results in R.


```{r}
file_path <- "data/Calculated Field Examples.xlsx"

sales <- read_excel(file_path, sheet = "Sales") %>% clean_names()
customers <- read_excel(file_path, sheet = "Customers") %>% clean_names()
products <- read_excel(file_path, sheet = "Product") %>% clean_names()
suppliers <- read_excel(file_path, sheet = "supplier") %>% clean_names()
```


```{r}
sales <- sales %>%
  mutate(
    date = as.Date(date),
    revenue = quantity * price * (1 - discount)
  )
```



```{r}
con <- dbConnect(SQLite(), "retail_database.sqlite")

dbWriteTable(con, "sales", sales, overwrite = TRUE)
dbWriteTable(con, "customers", customers, overwrite = TRUE)
dbWriteTable(con, "products", products, overwrite = TRUE)
dbWriteTable(con, "suppliers", suppliers, overwrite = TRUE)
```

## Total Revenue

```{r}
total_revenue <- dbGetQuery(con, "
SELECT SUM(quantity * price * (1 - discount)) AS total_revenue
FROM sales
")

knitr::kable(total_revenue)
```

## Top 10 Customers by Revenue

```{r}
top_customers <- dbGetQuery(con, "
SELECT
    c.name,
    SUM(s.quantity * s.price * (1 - s.discount)) AS total_revenue
FROM sales s
JOIN customers c
    ON s.customer_id = c.customer_id
GROUP BY c.name
ORDER BY total_revenue DESC
LIMIT 10
")

knitr::kable(top_customers)
```

## Top 10 Products by Revenue

```{r}
top_products <- dbGetQuery(con, "
SELECT
    p.product_name,
    SUM(s.quantity * s.price * (1 - s.discount)) AS total_revenue
FROM sales s
JOIN products p
    ON s.product_id = p.product_id
GROUP BY p.product_name
ORDER BY total_revenue DESC
LIMIT 10
")

knitr::kable(top_products)
```

## Revenue by Category

```{r}
category_revenue <- dbGetQuery(con, "
SELECT
    p.category,
    SUM(s.quantity * s.price * (1 - s.discount)) AS total_revenue
FROM sales s
JOIN products p
    ON s.product_id = p.product_id
GROUP BY p.category
ORDER BY total_revenue DESC
")

knitr::kable(category_revenue)
```

## Revenue by Region

```{r}
region_revenue <- dbGetQuery(con, "
SELECT
    c.region,
    SUM(s.quantity * s.price * (1 - s.discount)) AS total_revenue
FROM sales s
JOIN customers c
    ON s.customer_id = c.customer_id
GROUP BY c.region
ORDER BY total_revenue DESC
")

knitr::kable(region_revenue)
```

## Supplier Performance

```{r}
supplier_revenue <- dbGetQuery(con, "
SELECT
    sup.supplier_name,
    SUM(s.quantity * s.price * (1 - s.discount)) AS total_revenue
FROM sales s
JOIN products p
    ON s.product_id = p.product_id
JOIN suppliers sup
    ON p.supplier_id = sup.supplier_id
GROUP BY sup.supplier_name
ORDER BY total_revenue DESC
")

knitr::kable(supplier_revenue)
```

## Monthly Sales Trend

```{r}
monthly_sales <- sales %>%
  mutate(month = floor_date(date, "month")) %>%
  group_by(month) %>%
  summarise(total_revenue = sum(revenue, na.rm = TRUE), .groups = "drop")

knitr::kable(monthly_sales)
```

## Visualization: Top Customers

```{r}
ggplot(top_customers,
       aes(x = reorder(name, total_revenue),
           y = total_revenue,
           fill = name)) +
  geom_col() +
  coord_flip() +
  scale_fill_brewer(palette = "Set1") +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "Top 10 Customers by Revenue",
    x = "Customer",
    y = "Revenue"
  ) +
  theme(legend.position = "none")
```

## Visualization: Top Products

```{r}
ggplot(top_products,
       aes(x = reorder(product_name, total_revenue),
           y = total_revenue,
           fill = product_name)) +
  geom_col() +
  coord_flip() +
  scale_fill_brewer(palette = "Set1") +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "Top 10 Products by Revenue",
    x = "Product",
    y = "Revenue"
  ) +
  theme(legend.position = "none")
```

## Visualization: Revenue by Category

```{r}
ggplot(category_revenue,
       aes(x = reorder(category, total_revenue),
           y = total_revenue,
           fill = category)) +
  geom_col() +
  coord_flip() +
  scale_fill_brewer(palette = "Set1") +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "Revenue by Product Category",
    x = "Category",
    y = "Revenue"
  ) +
  theme(legend.position = "none")
```

## Visualization: Monthly Sales Trend

```{r}
ggplot(monthly_sales, aes(x = month, y = total_revenue, group = 1)) +
  geom_line(linewidth = 1.2, color = "#FC8D62") +
  geom_point(size = 2.5, color = "#FC8D62") +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "Monthly Sales Trend",
    x = "Month",
    y = "Revenue"
  ) 
```

## Key Findings

A small number of customers generate a large share of total revenue. Some products and categories also contribute more to sales than others. When the data is grouped by region, certain regions consistently produce higher revenue. The monthly sales trend shows how revenue changes over time and helps identify periods where sales increase or slow down.

## Conclusion

This project uses SQL and R to analyze retail sales data. SQL was used to join the tables and calculate revenue across customers, products, categories, and regions. R was used to clean the data and create visualizations that make the results easier to see. Overall, the analysis provides a clearer picture of where revenue is coming from and how sales change over time.

```{r}
dbDisconnect(con)
```